---
title: "Project3"
author: "Jung-Han Wang"
date: "Monday, November 17, 2014"
output: pdf_document
---

## Project 3

# Problem 1

This problem is to get some codes to perform the support vector data description (SVDD)

a. Write an _R_ function to perform the SVDD.
```{r}

```

b. Write an _R_ function to perform the prediction of a new observation using SVDD.
```{r}
```

c. Write an _R_ function for detecting potential outliers for a new set of oservations, along with the upper threshold.
```{r}
```


# Problem 2

The goal of problem 2 is to perform the support vector data description (SVDD) using the Mahalanobis kernel function. We will simplify the problem
by using the identity function for g.

(a) Write an _R_ function to compute the Mahalanobis kernel distance $\pmb{d_g(x)}$

(b) Write an _R_ function to perform the Mahalanobis kernel SVDD.

(c) Write an _R_ function to perform the prediction of a new observation using the Mahalanobis kernel SVDD.

(d) Write an _R_ function for detecting potential outliers for a new set of oservations, along with the upper threshold.


```{r,}

```


