---
title: "Project3"
author: "Robert Norberg, Jung-Han Wang"
date: "Monday, November 20, 2014"
output: pdf_document
---
```{r,echo=FALSE}
# Clear working environment
rm(list=ls())

## Load Package
library(quadprog)
```
## Project 3

# Problem 1

This problem is to get some codes to perform the support vector data description (SVDD)
```{r}
```


a. Write an _R_ function to perform the SVDD.

Read data training dataset in environment.
```{r}
##Load Data Wang's Path
setwd("E:/Cloud Storage/Dropbox/Life long study/Ph.D/Lecture/2014 Fall/
      Statistical Computing/Project 2/Project2/Project3")
##Load Data Your Path
setwd("")

train<-read.table("training dataset.txt")

```

b. Write an _R_ function to perform the prediction of a new observation using SVDD.
```{r}
```

c. Write an _R_ function for detecting potential outliers for a new set of oservations, along with the upper threshold.
```{r}
```


# Problem 2

The goal of problem 2 is to perform the support vector data description (SVDD) using the Mahalanobis kernel function. We will simplify the problem
by using the identity function for g.

(a) Write an _R_ function to compute the Mahalanobis kernel distance $\pmb{d_g(x)}$

(b) Write an _R_ function to perform the Mahalanobis kernel SVDD.

(c) Write an _R_ function to perform the prediction of a new observation using the Mahalanobis kernel SVDD.

(d) Write an _R_ function for detecting potential outliers for a new set of oservations, along with the upper threshold.


```{r,}

```


## Appendix with R code

```{r all-code, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE, tidy.opts=list(keep.blank.line=T)}
```


